\chapter{Material and method}
\label{chap:m&m}

\section{Material}
BD Accuri C6 Plus benchtop flow cytometer (BD Nordics (prev. Puls Norway), Norway). Filters and lasers.
CytoSub submersible flow cytometer (CytoBuoy, Netherlands)
Coulter Counter Multisizer4 (Beckman Coulter, US) eqipped with a 100 \micro m aperture (size-range 2-60 \micro m)

Nikon Eclipse Ni-U
- Semrock Brightline GFP-4050B filter-cube
- Semrock Brightline LED-Cy5-A
- Plan Fluor 40x/0.75 water immersion objective,
- Plan Fluor 100x/1.30 Oil immersion objective

Nikon Eclipse 90i
- Filter specs?
- Nikon Plan Apo 60XA/1.40 Oil immersion objectice
- Nikon Plan Apo VC 100X/1.40 Oil immersion objective

Leitz Labrolux 12 binocular microscope (Leica Mikroskopi AS, Norway) [EF 40/0.65 objective]
Eppendorf Centrifuge 5804 R (Eppendorf, Norway), [rotor: A-4-44, rotor radius: 15.5 cm], high-speed refrigerated benchtop centrifuge
Jouan KR22i floor centrifuge (Thermo Fischer Scientific, US), high-speed high capacity refrigerated floor centrifuge 8 [rotor: AK 100-21] angle?
Bürker Counting Chamber (Hirschmann Laborgeräte, Germany) with 0.1 mm depth of chamber
Eirik Lund sitt kamera, lense og imaging software: 
Sony ILCE A6400 with E-mount, lense: Tamron 17-70mm F/2.8 
Adobe\textsuperscript{\textregistered} Lightroom Classic 12.0 
BD Microlance$^{TM}$ 3, 23G x 1" - Nr. 16, 0.6mm x 25 mm, sterile, REF 300800
HENKE-JECT 1 mL syringes, 1mL, luer, REF 8300014579, HENKE SASS WOLF GmbH, Tuttlingen, Germany

\subsection{Laboratory instruments}
\begin{table}[H]
	\centering
	%\caption{Chemicals used in the master thesis, listed alphabetically according to chemical name, including the chemical's CAS nr., purity/grade, supplier and state.}
	\label{tb:instruments}
	\resizebox{\linewidth}{!}{
	\begin{tabular}{lll}
	\textbf{Instrument} & \textbf{Model} & \textbf{Producer} \\
		\midrule
   Benchtop Flow Cytometer               & BD Accuri$^{TM}$ C6 Plus & BD Biosciences \\
   Submersible Flow Cytometer            & Cytosub                  & CytoBuoy \\
   Coulter Counter                       & Multisizer 4             & Beckman Coulter \\
   Upright microscope                    & Eclipse Ni-U             & Nikon \\
   Upright microscope                    & Ecliplse 90i             & Nikon \\
   Transmitted/incident light microscope & Labrolux 12              & Leitz \\
   Light engine                          & Sola SM II 365           & Lumencor \\
   Light engine                          & EL6000                   & Leica Microsystems \\
    CMOS camera                           & MC170HD                  & Leica Microsystems \\
   Microscope camera                      & DS-Fi1                   & Nikon \\
   Altin's fluorescence camera           & ?                        & Nikon \\
   Microscope camera                     & 4KHDMI                   & DeltaPix \\
   Benchtop centrifuge                   & Centrifuge 5804 R        & Eppendorf\\
   Floor centrifuge                      & KR22i                    & Jouan \\
   Counting chamber                      & Bürker                   & Hirschmann \\
   		\bottomrule
	\end{tabular}
}
\end{table}



\subsection{Chemicals}
\begin{table}[H]
	\centering
	%\caption{Chemicals used in the master thesis, listed alphabetically according to chemical name, including the chemical's CAS nr., purity/grade, supplier and state.}
	\label{tb:chemical-list}
	\resizebox{\linewidth}{!}{
	\begin{tabular}{lllll}
	\textbf{Chemicals (abbrv.)} & \textbf{CAS-No.} & \textbf{Purity/grade} & \textbf{Supplier} & \textbf{state} \\
		\midrule
    Calcium chloride dihydrate      & 10035-04-8 & $\geq$ 99.0 \% & Sigma Aldrich & s \\
    Dimethyl sulfoxide              & 67-68-5    & $\geq$ 99.5 \% & Sigma Aldrich & l \\
    D-(+)-Glucose                   & 50-99-7    & $\geq$ 99.5    & Sigma Aldrich & s \\
    \ce{Na2EDTA}$\cdot$\ce{2H2O}    & 6381-92-6  & 98.5-101.5 \%  & Sigma Aldrich & s \\
    Ethanol                         & 64-17-5    & 96 \% vol      & VWR           & l \\
    \ce{Na2HPO4}$\cdot$\ce{2H2O}    & 10028-24-7 & $\geq$ 98.0    & Sigma Aldrich & s \\
    Potassium phosphate monobasic   & 7778-77-0  & $\geq$ 98.0    & Sigma Aldrich & s \\
    Copper(II)sulfate pentahydrate  & 7758-99-8  & $\geq$ 98.0    & Sigma Aldrich & s \\
    Formaldehyde                    & 50-00-0    & 37\% wt        & Sigma Aldrich & l \\
    HEPES                           & 7365-45-9  & $\geq$ 99.5 \% & Sigma Aldrich & s \\
    Magnesium sulfate heptahydrate  & 10034-99-8 & $\geq$ 99.5 \% & Sigma Aldrich & s \\
    Methanol                        & 67-56-1    & $\geq$ 99.9 \% & Sigma Aldrich & l \\
    Potassium chloride              & 7447-40-7  & $\geq$ 99.9 \% & Sigma Aldrich & s \\
    Sodium chloride                 & 7647-14-5  & $\geq$ 99.5 \% & Merck         & s \\
    Trizma\textsuperscript{\textregistered}base & 77-86-1    & ACS reagent    & Merck         & s \\
    TRIS HCl                        & 1185-53-1  & $\geq$ 99.0 \% & Sigma Aldrich & s \\
		\bottomrule
	\end{tabular}
	}
\end{table}


\subsection{Reagents for Flow Cytometry}
\begin{table}[H]
	\centering
	%\caption{Reagents and kits used in the master thesis, listed alphabetically according to product name, including manufacturer, supplier and supplier's catalogue number.}
	\label{tb:reagent-list}
	\resizebox{\linewidth}{!}{
	\begin{tabular}{lllll}
	\textbf{Product name (abbrv.)} & \textbf{Manufacturer} & \textbf{Supplier} & \textbf{Catalogue} & \textbf{Concentration} \\
		\midrule
    TO-PRO$^{TM}$-3 Iodide (642/661) &  InVitrogen$^{TM}$  & Thermo Fisher & T3605 & 1.2 \micro M \\
    Ethidium Homodimer-1 &  InVitrogen$^{TM}$ & Thermo Fisher &  E1169 & 4 \micro L/sample \\
    Apotracker$^{TM}$ Green & BioLegend & Fisher Scientific & 50-207-9934 & 560 nM \\
    Calcein-AM & Invitrogen$^{TM}$ & Thermo Fisher & C1430 & 170 nM \\ 
    CS\&T RUO beads & BD Biosciences & BD Biosciences & 661414 &  4 drops/mL \\
    8-peak validation beads & Spherotech & BD Biosciences & 653144 & 4 drops/mL \\
    6-peak validation beads & Spherotech & BD Biosciences & 653145 & 4 drops/mL \\
		\bottomrule
	\end{tabular}
	}
\end{table}


\subsection{Microscopy kits and reagents}
\begin{table}[H]
	\centering
	%\caption{Reagents and kits used in the master thesis, listed alphabetically according to product name, including manufacturer, supplier and supplier's catalogue number.}
	\label{tb:Microscopy-list}
	\resizebox{\linewidth}{!}{
	\begin{tabular}{llll}
	\textbf{Product name (abbrv.)} & \textbf{Producer} & \textbf{Supplier} & \textbf{Catalogue} \\
		\midrule
    Giemsa's azur eosin methylene blue solution & Merck & Sigma Aldrich & 1.09204.0500 \\
    Hemacolor\textsuperscript{\textregistered} & Sigma Aldrich & Sigma Aldrich & 1.11661 \\
    Eukitt\textsuperscript{\textregistered} Quick-hardening mounting medium & Orsatec GmbH & Sigma Aldrich & 03989 \\
    Type N Immersion Oil for Microscopy & Nikon & ? & MXA20234 \\
    Percoll$^{TM}$ & Cytiva Sweden AB & Sigma Aldrich & GE17-0891-02 \\
		\bottomrule
	\end{tabular}
	}
\end{table}



\subsection{Buffers and solutions}
\begin{table}[H]
	\centering
	\label{tb:buffers}
	\resizebox{\linewidth}{!}{
	\begin{tabular}{ll}
	\textbf{Buffer} & \textbf{Composition} \\
		\midrule
    MAS                   &  375.6 mM \ce{NaCl}, 28.97 mM Citric Acid$\cdot$3Na$\cdot$2\ce{H2O}, 113.8 mM D-Glucose, \\ 
                          & 2.6 mM Citric Acid$\cdot$\ce{H2O}, 11.5 mM \ce{Na2EDTA}$\cdot$\ce{2H2O}, pH=7.0, 0.2 \micro m filtered \\
    Anticoagulant buffer  & 55.5 mM D-glucose, 171.1 mM NaCl, 13.4 mM \ce{Na2EDTA}$\cdot$\ce{2H2O}, \\
                          & 0.05 M TRIS/HCl, pH=7.6, 0.2 \micro m filtered \\ 
    PBS                                  & 136.9 mM \ce{NaCl}, 2.7 mM \ce{KCl}, 10.1 mM \ce{Na2HPO4}, 1.8 mM \ce{KH2PO4} \\
    Sorensen Buffer       & 66.7 mM \ce{KH2PO4}, 66.7 mM \ce{Na2HPO4}$\cdot$\ce{2H2O}, pH=6.8, 0.2 \micro m filtered \\
    MPSS    &  470 mM \ce{NaCl}, 10 mM \ce{KCl}, 10 mM \ce{CaCl2}, 10 mM HEPES \\
                          & 47.7 mM \ce{MgSO4}, pH=7.41, 0.2 \micro m filtered \\
    Tris Buffered Saline  & 44.5 mM Trizma\textsuperscript{\textregistered}base, 5.5 mM TRIS-HCl, 450 mM \ce{NaCl}, pH=7.00 \\
		\bottomrule
	\end{tabular}
	}
\end{table}

\section{Method}
\subsection{Animal housing}
Adult blue mussels (\emph{Mytilus edulis}) of x.x$\pm{5}$ cm shell length were obtained from Snadder og Snaskum AS (Indre Fosen, Norway). Upon arrival at the marine animal housing facilities of NTNU, Centre of Fisheries and Aquaculture (SeaLab), the mussels were transferred to 50 L filtered seawater flow-through tanks (11 L/min) supplied by a direct inlet from Trondheimsfjorden at 80 m depth ($\SI{7.5}{\celsius}$). Here, the mussels were kept to acclimatize for 2 days before transfer to the experimental exposure setup.

Time of mussel harvest/purchase?

Describe water treatment in more detail: include information regarding sand-filter, protein-skimmer, 0.5 um filter bags and UV-treatment.

Because the watertreatment left no natural feed for the mussels, the filtered seawater were suplemented with algea (species) [frequency]


\subsection{Experimental setup/design}
- Specs related to flowrate, feeding (alge conc: Coulter counter), how often was nano particles changed (every 3-4 days: check article), concentrations (ICP-MS --> mg/L in stocks and exp tanks), zetasizer --> size and surface potential

include figure of exp setup?

depuration period: check article

storage from depuration to sampling: mussels were not kept on ice, but were washed and taken directly upstairs for measuring weight, length, height and width for condition index, and were given to us for FCM sampling afterwards. If we were delayed, they were kept in the fridge until hemolymph sampling 


\subsection{Hemolymph collection}
By employing the sampling technique described above, 0.5 mL hemolymph was withdrawn from each mussel into 1.0 mL syringes pre-filled with 0.5 mL anticoagulant buffer (ACB).

- Hemolymph was not pooled or centrifuged. 

\subsection{Determination of hemocyte concentration and subpopulations}
"The total hemocyte concentration, morphometry and definition of sub-populations were determined using side scatter (SSC) and forward scatter (FSC) light. Side scatter light corresponds to the relative internal complexity or granularity of the cells and FSC corresponds to the relative cell size" \cite{Rolton2020}

Concider including the experiment that was performed to validate the BD Accuri C6 Plus hemocyte counting technique with a technique using counting beads. Mention that a hemocytometer/counting chamber was used for the initial method development. It's a lot of data and work, which never looks bad. [Maybe not a separate subsection, but include under FCM or elsewere?]

\subsection{Cytologic characterization of M.edulis hemocyte subpopulations}
\label{subsection:morph}
For the purpose of characterizing and imaging the different haemocyte cell types of \emph{M. edulis} in a non-spread state, samples were withdrawn into an equal volume of ice-cold MPSS and transferred directly onto glass slides. The glass slides were kept in a humid chamber for up to 5 minutes before cells were fixed in ice cold methanol (5$\times$ 1 sec dips). This assured haemocyte attachment, but circumvented the profound morphological changes accompanied by the haemocytes’ process of spreading. Slides were stained with the Hemacolor\textsuperscript{\textregistered} kit according to the manufacturer’s recommendations, air dried, mounted with Eukitt\textsuperscript{\textregistered} and coverslipped. Stained haemocytes were examined and imaged under brightfield illumination by a $\times$100 oil immersion objective with 1.40 numerical aperture and a Nikon DS-Fi1 microscope camera.

Haemocyte morphology was primarily characterized on the basis of cytoplasmic staining, granularity (granule size, abundance and staining affinities), nuclear:cytoplasmic (N:C) ratio, size and shape. Since Le Foll and colleagues (2010) reported that the motile properties of the different haemocyte subpopulations of \emph{M. edulis} could be used as an additional functional criteria for their characterization, the spreading behavior of the different cell types were included to expand on the morphological characteristics. This comprehended a description of the typical haemocyte shapes and motile structures that were observable under differential interference contrast illumination in spread haemocytes. The preparation of slides with spread haemocytes followed a methodology similar to that of the non-spread haemocytes, but the humid chamber incubation periods prior to fixation and staining were extended (15-30 min). Microscopic fields were examined through a $\times$60 oil immersion objective with 1.40 numerical aperture.

\subsection{Haemocyte differential and determination of size distributions}
\label{subsection:CytCar}
To determine the relative proportions (\%) of cell types in the haemolymph of untreated mussels, haemolymph from 20 adult mussels (shell length 55$\pm{3}$ mm) were withdrawn into an equal volume of 5\% formaldehyde in MPSS, and fixed for one hour in suspension. Fixed cells were pelleted by centrifugation (250G, 15 min, \SI{10}{\celsius}) and resuspended in 100 \micro L 0.75 \% eosin in Sorensen Buffer. After staining for 5 minutes, 1 mL 3\% Wright’s-Giemsa was added and the staining was continued for another 15 minutes. Stained haemocytes were pelleted by centrifugation (180G, 12 min, \SI{10}{\celsius}), resuspended in 100-200 \micro L Sorensen Buffer and transferred onto glass slides. The slides were air-dried in a fume hood until the smears were completely dry and transparent, before they were mounted with Eukitt\textsuperscript{\textregistered} and coverslipped.

A differential count was performed by counting a minimum of 1000 haemocytes from each of the 20 untreated mussels. On the basis of the morphological criteria in section \ref{subsection:morph}, the haemocytes where placed in one of three categories: (1) eosinophilic granulocytes, (2) basophilic granulocytes or (3) small blast-like basophils. To establish wether the three cell types differed with regard to size across a larger population of adult mussels, the differential was combined with cell diameter measurements. Throughout each differential, 10-20 microscopic fields were imaged under DIC illumination by a $\times$60 oil immersion objective, and the diameters of 100 haemocytes were measured using the straight line tool of ImageJ image processing software (NIH, Bethesda, Maryland, US). The scale was set with a stage micrometer calibration slide (Leica Microsystems, Buffalo, US). When haemocyte outlines deviated from an approximate spherical shape, the diameter was estimated as the average length of the long and short axes, after Burkhard et al. (2009).

\subsection{Flow Cytometry}
Flow Cytometer used and Flow Cytometry acquisition software
External software used for graphing and analysis of exported FCS files.
Replicate/triplicate measurements?
Number of events recorded for each mussel?
Briefly describe FCS and SSC.
Mention if data were collected in linear or logarithmic scale, 
FSC treshold (80.000 FSC-H). 
Fluorescent compensation (matrix)
Describe gating strategy herein? Debris exclusion (size); doublet exclusion --> hemocytes. (Gating of basophils and eosinophils, the use of FMO controls with TO-PRO-3 and/or the apoptosis stain) to create gates.


\begin{table}[H]
	\centering
	\caption{The FCM acquisition and fluidics settings specified with the BD Accuri C6 Plus acquisition software during the flow cytometric experiments reported in this work.}
	\label{tb:FCM_settings}
	\resizebox{\linewidth}{!}{
	\begin{tabular}{lllll}
	\textbf{Experiment nr.} & \textbf{Event-triggering threshold} & \textbf{Acquisition stop-condition} & \textbf{Flow rate (\micro L/min)} & \textbf{Core size (\micro m)} \\
		\midrule
    Aggregation & 80.000 FSC-H & acquired volume, 20 \micro L & 30 & 10 \\
    Calcein AM and TO-PRO$^{TM}$-3 Iodide & 80.000 FSC-H & 10.000 hemocyte events & 36 & 16 \\
    Apo-15 and TO-PRO$^{TM}$-3 Iodide & 80.000 FSC-H & 10.000 hemocyte events & 36 & 16 \\
		\bottomrule
	\end{tabular}
	}
\end{table}

\subsection{Isopycnic centrifugation}
Formaldehyde-fixed haemocytes were separated on discontinuous Percoll gradients according to the protocol by Friebel and Renwrantz (1995), with minor modifications. The separation was performed in duplicate with pooled haemocytes from a total of six adult mussels (shell length 55-70 mm). For both gradients, the haemolymph of three individual mussels (1.5-2.0 mL/mussel) were withdrawn into an equal volume of 5\% formaldehyde in MPSS, wherein the haemocytes were fixed for one hour after pooling. The fixed haemocytes were stained with 0.75 \% Eosin and 3 \% Giemsa according to the procedure in section~\ref{subsection:CytCar}. After pelleting, stained haemocytes were resuspended in Tris-Buffered Saline (TBS, 900 mOsm) to an approximate concentration of $9\times10^{6}$ cells/mL, before layering 2 mL suspension on top of both gradients. Discontinuous Percoll gradients consisted of 15\%, 33\%, 38\%, 43\% and 90\% Percoll stock in TBS (vol/vol), and were constructed by carefully layering 2 mL of each Percoll concentration in 15 mL Falcon centrifuge tubes with conical bottoms (Corning, New York, US).

The centrifugation was started at 120G for 10 minutes (\SI{4}{\celsius}), followed by 40 minutes at 2500G. An Eppendorf 5804 R benchtop centrifuge equipped with an A-4-44 swing-bucket rotor was employed for the separation, with the brake ramp set to 1. The separated cell fractions were collected into syringes by puncturing the tubes below the gradient interfaces with 23G hypodermic needles. According to a protocol by Bachére and Grizel (1988), the Percoll was eliminated from the 43/90\% and 38/43\% fractions by two consecutive dilutions in TBS (1:7) and a 10\% sucrose cushions (1:1) before centrifugation (800G, 15 min, \SI{4}{\celsius}). Haemocytes collected from the 15/33\% interface were centrifuged directly after a seven-fold dilution in TBS (800G, 15 min, \SI{4}{\celsius}).

The pelleted cell fractions were resuspended in Sorensen buffer and divided into two aliquots; one aliquot were used to prepare smears according to the procedure in section \ref{subsection:CytCar}, while the remaining aliquots were further diluted in 1 mL Sorensen buffer for flow cytometric characterization. A total of 10.000 events were acquired from each cell fraction on the flow cytometer (36 \micro L flow rate, 16 \micro m core size), in addition to 30.000 events from both pools that were set aside prior to centrifugation. The relative proportion (\%) of cell types in each fraction were ascertained by performing 1000-cell differential counts.

\subsection{Slide preparation}
Refer to the MN cytome assay by Bolognesi, and mention that we fixed and stained hemocytes in suspension to produce slides without aggregated hemocytes. Since the cells were dead, they did not spread or produce pseudopodia, which allowed us to observe the actual size and form which they would posses ass they flowed through the laser of the flow cytometer.

\subsection{Assay Validation}
Both dead cell stain and apoptosis stain has to be validated. For TO-PRO-3 perform the same experiment with 70\% methanol-killed cells as you did for Calcein-AM + EthD-1, and use Calcein-AM to stain live cells.
How to do this for the Apo-stain?

\subsection{Measurements and calculations}
(\cite{R-project})